---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
In this document we present the exploratory analysis performed on the 
[Activity monitoring dataset](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) 
as a part of Peer Review Assesment for the *Reproducible Research* course on Coursera. 

```{r load_data}
activity <- read.csv(unz("activity.zip", "activity.csv"), 
                     header = T, 
                     colClasses = c("integer", "character", "integer"))
str(activity)
```


## What is mean total number of steps taken per day?
We will ignore the missing values initially.
```{r removeNA}
# Remove missing values
activity_nonNA <- subset(activity, is.na(steps) == F)
```

Calculate the total number of steps taken each day
```{r mean_total}
daily_totals <- tapply(activity_nonNA$steps, activity_nonNA$date, sum)
print(daily_totals)
```

Plot the histogram of total number of steps taken each day.
```{r hist1, fig.height=4, fig.width=6}
hist(daily_totals, main = "Histogram of total number of steps taken each day", 
     xlab = "Steps")
```

Here are the **mean** and **median** total steps taken per day.
```{r steps_summary}
# Compute mean, median and save them into a dataframe
summarySteps <- data.frame(mean = mean(daily_totals), 
                           median = median(daily_totals))
summarySteps
```
 

## What is the average daily activity pattern?
Calculate the average number of steps taken in an interval, averaged across all 
days.
```{r activity_avgSteps}
avgStepsPerInterval <- tapply(activity_nonNA$steps, 
                              activity_nonNA$interval, mean)
head(avgStepsPerInterval)
```

Plot of 5-minute interval and the average number of steps in that interval across all days.
```{r ts}
plot(as.integer(names(avgStepsPerInterval)), avgStepsPerInterval, type = "l", 
     main = "Average number of steps per interval",
     xlab = "Interval", ylab = "Average number of steps")
```

Find out which 5-minute interval contains the maximum number of steps. 
```{r maxSteps}
names(which.max(avgStepsPerInterval))
```



## Imputing missing values
Here's the total number of missing values in the dataset. 
```{r numNA}
length(which(is.na(activity$steps) == T))
```

In order to fill up missing values, we will use the mean for that 5-minute interval. We will create a new data set with the missing values filled up. For this, we will use *avgStepsPerInterval* calculated in the previous section.
```{r fillNA}
activityFilled <- activity
for (i in 1:nrow(activityFilled)) {
        if (is.na(activityFilled[i, 1])) {
                index <- as.character(activityFilled[i, 3])
                activityFilled[i, 1] <- avgStepsPerInterval[index]             
        }
}
head(activityFilled)
```

Here's a histogram of mean total number of steps taken each day (with missing values filled in).
```{r mean_total_new, fig.height=4, fig.width=6}
daily_totals_new <- tapply(activityFilled$steps, activityFilled$date, sum)
hist(daily_totals_new, main = "Histogram of total number of steps taken each day", 
     xlab = "Steps")
```

The **mean** and **median** values are as folows.
```{r}
mean(daily_totals_new)
median(daily_totals_new)
```

Let's compare the mean and median between the datasets without missing values and the one with missing values filled in.
```{r steps_summary_new}
summarySteps <- rbind(summarySteps, c(mean(daily_totals_new), 
                                      median(daily_totals_new)))
rownames(summarySteps) <- c("NA-ignored", "NA-filled")
summarySteps
```

They are same.

## Are there differences in activity patterns between weekdays and weekends?
First, let's find out whether the given observation belongs to a weekday or a weekend.
```{r week}
# Find day of the week
days <- weekdays(strptime(activityFilled$date, format = "%F"))

# convert day of the week into a "weekday" or a "weekend"
weekends <- which(days %in% c("Saturday", "Sunday"))
days[weekends] <- "weekend"
days[-weekends] <- "weekday"
days <- as.factor(days)
activityFilled <- cbind(activityFilled, days)
```

Let's compare average number of steps taken for each 5-minute interval on weekdays and weekends.
```{r plotPanel}
library(lattice)
library(reshape2)
# Compute interval means for each type of day
iM <- tapply(activityFilled$steps, 
             list(activityFilled$interval, activityFilled$days), 
             mean)
# Reshape the dataframe so that we can plot it
iMeans <- melt(iM)
names(iMeans) <- c("interval", "day", "steps")
xyplot(steps ~ interval | day, data = iMeans, type = "l", xlab = "Interval", 
       ylab = "Average number of steps", 
       main = "Average number of steps per interval (Weekdays vs. Weekends)",
       layout = c(1, 2),
       panel = function(x, y, ...){
               panel.xyplot(x, y, ...)
               # Add a horizontal line at the mean
               panel.abline(h = mean(y), lty = 2)
       })
```
